Implementation and assessment of tools for integrating forest roads into projections of the cumulative effects of disturbance on wildlife

### 2024-05-28 analysis/paper/supplement/supplement.Rmd
Updates for Forest Science

---
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      number_sections: yes
bibliography: Roads.bib
csl: "../templates/canadian-journal-of-forest-research.csl" # Insert path for the bib-style

@@

# Grade penalty function

@@

# Simple cost results

The simple cost method for setting edge weights generally performed worse than the grade penalty and Hardy methods (compare Figures 2 and \@ref(fig:aspat-perf-fig-simp), and Figures 5 and \@ref(fig:spat-perf-fig-simp)).

```{=html}
<!-- To remove? This is all covered in the main text and I don't think differs much here.
For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks, outside cutblocks, or overall (Figure \@ref(fig:aspat-perf-fig-simp)). Within all mapped cutblocks (Figure \@ref(fig:aspat-perf-fig-all)), the Hardy method most accurately projected road density, and least accurately projected distance to road. Low density method variants most severely underprojected road density within cutblocks, and most accurately projected distance to road within cutblocks (Figure \@ref(fig:aspat-perf-fig-all)). These apparently contradictory results are explained by areas in our input data that have been misidentified as cutblocks with no associated observed roads (e.g. Figure \@ref(fig:proj-map-fig)b). Omitting cutblocks without observed roads from the analysis (Figure \@ref(fig:aspat-perf-fig-simp)) clarifies that within real cutblocks the Hardy method most accurately projected road density and distance to road. However, note that our observed landscape is densely roaded in part because switchbacks are required in steep terrain , and the Hardy method projects a distinctly different pattern of road tortuosity within cutblocks (Figure \@ref(fig:proj-meth-fig)). None of the simple cost methods considered here can project switchbacks (Figure \@ref(fig:proj-meth-fig)).

Another fundamental limitation of these methods is that they generate straight road segments within pixels. As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the cost raster resolution is coarse (Figure A-\@ref(fig:figA1)). In contrast, projections of the proportion of cells that contain roads is more accurate on a coarse raster (Figure A-\@ref(fig:figA1)).

Outside cutblocks, all method variants underproject road density and road presence, and there is little variation in performance among them (Figure \@ref(fig:aspat-perf-fig-simp)). Comparison of observed and projected road networks clarifies that projected roads cross valley bottoms far more often than observed roads which often run parallel to one another on each side of rivers (Figure \@ref(fig:proj-map-fig)a). Accounting for the costs of crossing streams, rivers and other wet areas in valley bottoms would likely reduce this discrepancy. Switchbacks on steep slopes (Figure \@ref(fig:proj-map-fig)c) are also a source of discrepancy outside of cutblocks. Overall, projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint are more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig-simp)), suggesting that these measures are easier to project. Performance differences among low and high density method variants are larger than differences among projection algorithms, and all road network projection method variants represent an improvement over the "cutblocks only" scenario where cutblocks were the only disturbance considered.
-->
```
```{r get-data}
# load csv
meanTable <- read.csv(here(data_path_drvd,"../TSA27", "mean_table.csv"))
matchData <- read.csv(here(data_path_drvd,"../TSA27", "agree_table.csv"))
meanTableRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts",
                                   "mean_table.csv"))

@@

# All cutblocks results

@@

# Coarse resolution results

@@

# Benchmarking road projection methods

@@

Execution time and memory requirements increase with the number of nodes in the graph, and the slope of that increase is highest for the MST method (Figure \@ref(fig:speed-fig)). Except on the very small landscape the ILCP method is faster than the MST method. Reducing the number of landings by targeting centroids also reduces execution time (Figure \@ref(fig:speed-fig)). The Hardy method is significantly faster than any of the R package methods.

@@

@@

minor_breaks <- rep(1:4*2, 21)*(10^rep(-10:10, each=9))
bench_res %>%
  add_row(cutblocks_real = "revelstoke_real", sampleType = "regular", sampleDens = 1.5e-05, 
          method = "Hardy QGIS", Elapsed_Time_sec = hardy_time, n_verticies = 1815825) %>% 
  filter(cutblocks_real == "revelstoke_real", sampleType %in% c("regular", "centroid"), 
         sampleDens > 1e-06 | sampleType == "centroid") %>% 
  mutate(Peak_RAM_Used_MiB = Peak_RAM_Used_MiB * 0.001048576, 
         sampleDens = ifelse(sampleType == "centroid", 0.1, sampleDens * 1000^2)) %>% 
  pivot_longer(c(Elapsed_Time_sec, Peak_RAM_Used_MiB), names_to = "variable", 
               values_to = "values") %>% 
  ggplot(aes(y = values, col = interaction(method, sampleDens), x = n_verticies,
             group = interaction(method, sampleDens), 
             shape = interaction(method, sampleDens)))+
  geom_line(linewidth = 1.05)+
  geom_point(size = 2)+
  scale_colour_manual(labels = c("ILCP centroid", "MST centroid",
                                 "ILCP high density", "MST high density", "Hardy QGIS"),
                      values = c("#ab84a5","#732f30", "#df8d71", "#1e5a46", "#af4f2f"))+
  scale_shape_discrete(solid = FALSE, 
                       labels = c("ILCP centroid", "MST centroid", 
                                  "ILCP high density", "MST high density", "Hardy QGIS"))+
  scale_y_log10(minor_breaks = minor_breaks)+
  scale_x_log10(minor_breaks = minor_breaks, limits = c(1e04, NA))+
  facet_wrap(~variable, nrow = 2, scales = "free", 
             labeller = labeller(variable = c(Elapsed_Time_sec = "Time (s)", 
                                              Peak_RAM_Used_MiB = "Peak RAM used (GB)"))) +
  labs(y = NULL, x = "# Nodes", col = "Method", shape = "Method")+
  theme_bw()
```

# Profiling

Profiling with the `Rprof` function shows which parts of a function take the longest. On a test run with 1 million cells in the weight raster and 10 landings we profiled the performance of the ILCP and MST methods. Both methods begin by constructing a full graph, which took 3-5 s. For the ILCP method, iteratively determining the shortest paths is the only other step that takes more than a second to run, and this run time will increase with the number of landings (Figure \@ref(fig:profiling)). For the MST method, calculating cost distances among landings and from each landing to the nearest point on the existing road takes the most time, which also increases with the number of landings (Figure \@ref(fig:profiling)). Determining the locations of the paths in the minimum spanning tree also takes about 6 seconds. We experimented with using graph construction methods in the package `gdistance` [@van_etten_r_2017], but Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] method is faster (Figure \@ref(fig:profiling)).

@@

# Fort Nelson map

@@

# References

@@

# Colophon

>>>>

---
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      number_sections: yes
bibliography: ../Roads.bib
csl: "../../templates/canadian-journal-of-forest-research.csl" # Insert path for the bib-style

@@

## Grade penalty function

@@

## Simple cost results

The simple cost method for setting edge weights generally performed worse than the grade penalty and Hardy methods (compare Figures 2 and S\@ref(fig:aspat-perf-fig-simp), and Figures 5 and S\@ref(fig:spat-perf-fig-simp)).

```{r get-data}
# load csv
meanTable <- read.csv(here(data_path_drvd,"../TSA27", "mean_table.csv"))
matchData <- read.csv(here(data_path_drvd,"../TSA27", "agree_table.csv"))
meanTableRealCuts <- read.csv(here(data_path_drvd,"../TSA27_real_cuts",
                                   "mean_table.csv"))

@@

## All cutblocks results

@@

## Coarse resolution results

@@

## Benchmarking road projection methods

@@

Execution time and memory requirements increase with the number of nodes in the graph, and the slope of that increase is highest for the MST method (Figure S\@ref(fig:speed-fig)). Except on the very small landscape the ILCP method is faster than the MST method. Reducing the number of landings by targeting centroids also reduces execution time (Figure S\@ref(fig:speed-fig)). The Hardy method is significantly faster than any of the R package methods.

@@

ch_time <- structure(list(id = 1, resolution = 50, n_verticies = 21791850, 
    n_edges = 87138977, Elapsed_Time_sec = 10001.606, Total_RAM_Used_MiB = 4018.7, 
    Peak_RAM_Used_MiB = 10752.7), class = "data.frame", row.names = c(NA, 
-1L))

@@

minor_breaks <- rep(1:4*2, 21)*(10^rep(-10:10, each=9))
bench_res %>%
  add_row(cutblocks_real = "revelstoke_real", sampleType = "regular", sampleDens = 1.5e-05, 
          method = "Hardy QGIS", Elapsed_Time_sec = hardy_time, n_verticies = 1815825) %>% 
  # add_row(cutblocks_real = "revelstoke_real", sampleType = "centroid", sampleDens = 1.5e-05, 
  #         method = "ilcp", Elapsed_Time_sec = ch_time$Elapsed_Time_sec, 
  #         n_verticies = ch_time$n_verticies, Peak_RAM_Used_MiB = ch_time$Peak_RAM_Used_MiB) %>% 
  filter(cutblocks_real == "revelstoke_real", sampleType %in% c("regular", "centroid"), 
         sampleDens > 1e-06 | sampleType == "centroid") %>% 
  mutate(Peak_RAM_Used_MiB = Peak_RAM_Used_MiB * 0.001048576, 
         sampleDens = ifelse(sampleType == "centroid", 0.1, sampleDens * 1000^2)) %>% 
  pivot_longer(c(Elapsed_Time_sec, Peak_RAM_Used_MiB), names_to = "variable", 
               values_to = "values") %>% 
  ggplot(aes(y = values, col = interaction(method, sampleDens), x = n_verticies,
             group = interaction(method, sampleDens), 
             shape = interaction(method, sampleDens)))+
  geom_line(linewidth = 1.05)+
  geom_point(size = 2)+
  scale_colour_manual(labels = c("ILCP centroid", "MST centroid",
                                 "ILCP high density", "MST high density", "Hardy QGIS"),
                      values = c("#ab84a5","#732f30", "#df8d71", "#1e5a46", "#af4f2f"))+
  scale_shape_discrete(solid = FALSE, 
                       labels = c("ILCP centroid", "MST centroid", 
                                  "ILCP high density", "MST high density", "Hardy QGIS"))+
  scale_y_log10(minor_breaks = minor_breaks)+
  scale_x_log10(minor_breaks = minor_breaks, limits = c(1e04, NA))+
  facet_wrap(~variable, nrow = 2, scales = "free", 
             labeller = labeller(variable = c(Elapsed_Time_sec = "Time (s)", 
                                              Peak_RAM_Used_MiB = "Peak RAM used (GB)"))) +
  labs(y = NULL, x = "# Nodes", col = "Method", shape = "Method")+
  theme_bw()
```

## Profiling

Profiling with the `Rprof` function shows which parts of a function take the longest. On a test run with 1 million cells in the weight raster and 10 landings we profiled the performance of the ILCP and MST methods. Both methods begin by constructing a full graph, which took 3-5 s. For the ILCP method, iteratively determining the shortest paths is the only other step that takes more than a second to run, and this run time will increase with the number of landings (Figure S\@ref(fig:profiling)). For the MST method, calculating cost distances among landings and from each landing to the nearest point on the existing road takes the most time, which also increases with the number of landings (Figure S\@ref(fig:profiling)). Determining the locations of the paths in the minimum spanning tree also takes about 6 seconds. We experimented with using graph construction methods in the package `gdistance` [@van_etten_r_2017], but Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] method is faster (Figure S\@ref(fig:profiling)).

@@

## Fort Nelson map

@@

# Literature Cited {.unnumbered}

@@

# Colophon {.unnumbered}
